The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-08-02 paper2_JNLS/paper.tex
NM

%----------------------------------------------------------------------------%
%% LATEX DEFINITIONS
%----------------------------------------------------------------------------%
% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}
\newcommand{\dx}{\, dx}
% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{\mathcal{D}}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\En}{\mathcal{E}}
\newcommand{\Mo}{\mathcal{M}}
\newcommand{\skw}{\text{skew}}
\newcommand{\skwdn}{\skw_+}
\newcommand{\var}{\text{var}}
\newcommand{\varup}{\var_-}
\newcommand{\vardn}{\var_+}
% Dimensionless parameters
\newcommand{\ampscale}{\mathcal{A}}
\newcommand{\lengthscale}{\mathcal{L}}
\newcommand{\timescale}{\mathcal{T}}
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian and Gibbs stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\sympJ}{\mathcal{J}}
\newcommand{\vard}[2]{\frac{\delta #1}{\delta #2}}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\Hthree}{\Ham_{3}}
\newcommand{\Htwo}{\Ham_{2}}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}

@@

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ is a piece-wise constant function of dimensionless time
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}

@@

%NOTE: Perhaps we should compare the dimensionless parameters to our PNAS paper \cite{majda2019statistical}, BUT the expressions for E0 and L0 given in the PNAS paper do not match with what I calculate comparing to the above. Plus, it seems like it could only add confusion.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\dot{u} = \sympJ \vard{\Ham}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham$ the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\Ham = C_2 \drat^{1/2} \, \Htwo - C_3 \drat^{-3/2} \, \Hthree
\end{equation}
where $\drat$ changers value across the ADC. More explicitly, the separate upstream and downstream Hamiltonians are given by
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree  \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With those signs, \cite{majda2019statistical} found  {\em negative} inverse temperature to be appropriate for recovering the experimental observations. We used the signs above so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional statistical mechanics.
% Notes for myself on the sign change: Abramov CPAM 2003 and Bajars

%----------------------------------------------------------------------------%
% Truncation
%----------------------------------------------------------------------------%
\subsection{The truncated KdV framework}
\label{tKdVSec}

>>>>

%----------------------------------------------------------------------------%
%% LATEX DEFINITIONS
%----------------------------------------------------------------------------%
% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}
\newcommand{\dx}{\, dx}
% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{\mathcal{D}}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\En}{\mathcal{E}}
\newcommand{\Mo}{\mathcal{M}}
\newcommand{\skw}{\text{skew}}
\newcommand{\skwdn}{\skw_+}
\newcommand{\var}{\text{var}}
\newcommand{\varup}{\var_-}
\newcommand{\vardn}{\var_+}
% Dimensionless parameters
\newcommand{\ampscale}{\mathcal{A}}
\newcommand{\lengthscale}{\mathcal{L}}
\newcommand{\timescale}{\mathcal{T}}
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\sympJ}{\mathcal{J}}
\newcommand{\vard}[2]{\frac{\delta #1}{\delta #2}}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\Hup}{\Ham_{-}}
\newcommand{\Hdn}{\Ham_{+}}
\newcommand{\Hupdn}{\Ham_{\pm}}
\newcommand{\Hthree}{\Ham_{3}}
\newcommand{\Htwo}{\Ham_{2}}
\newcommand{\Fcnl}{\mathcal{F}}
% Gibbs and theta
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}

@@

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ is a piece-wise constant function of dimensionless time
\begin{equation}
\label{dratpw}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}

@@

%NOTE: Perhaps we should be more specific about comparing to the dimensionless parameters in our PNAS paper \cite{majda2019statistical}, BUT the expressions for E0 and L0 given in the PNAS paper do not match with what I calculate comparing to the above. Plus, it seems like it could only add confusion.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\dot{u}_{\pm} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw}, the separate upstream and downstream Hamiltonians are
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With that sign convention, a {\em negative} inverse temperature is required to accurately describe the experimental observations, as found in \cite{majda2019statistical}. We, however, choose the sign convention above so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional form of statistical mechanics.

% Notes for myself on the sign change: Abramov CPAM 2003, Bajars Nonlinearity 2013, and our 2019 PNAS paper all use J = -d/dx with the opposite sign in the Hamiltonian. However, I was able to easily find several references online that use J = d/dx, so there is nothing wrong with it. The exact criteria that J has to satisfy are listed in the CPAM 2003 paper, and J = d/dx satisfies all these criteria just as well as J = -d/dx does.

The evolution of any functional $\Fcnl[u]$ is given by 
\begin{align}
\partial_t \Fcnl = \{ \Fcnl, \Ham \} = \int_{-\pi}^{\pi} \vard{F}{u} \sympJ \vard{\Ham}{u} \dx
\end{align}
where again $\Ham = \Hup$ if $t<0$ and $\Ham = \Hdn$ if $t>0$.
%----------------------------------------------------------------------------%
% Truncation
%----------------------------------------------------------------------------%
\subsection{The truncated KdV framework}
\label{tKdVSec}
